---
title: "Global rsofun run"
author: "Beni Stocker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{map2tidy functionality}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
source_files <- list.files(here::here("R/"), "*.R$")  # locate all .R files
purrr::map(paste0(here::here("R/"), source_files), ~source(.)) 
library(tidyverse)
library(rnaturalearth)
library(cowplot)
```

## Specify settings

```{r}
settings <- list(
  fileprefix = "test",
  model = "pmodel",
  year_start = 2018,  # xxx not yet handled
  year_end = 2018,  # xxx not yet handled
  grid = list(
    lon_start = -179.75,
    dlon = 0.5,
    len_ilon = 720,
    lat_start = -89.75,
    dlat = 0.5,
    len_ilat = 360
    ),
  source_climate = "watch-wfdei",
  dir_climate = "~/data/watch_wfdei",
  dir_climate_tidy = "~/data/watch_wfdei/tidy/",
  source_fapar = "modis",
  file_fapar = "~/data/MODIS-C006_MOD15A2_LAI_FPAR_zmaw/MODIS-C006_MOD15A2__LAI_FPAR__LPDAAC__GLOBAL_0.5degree__UHAM-ICDC__2000_2018__MON__fv0.02.nc",
  dir_fapar_tidy = "~/data/MODIS-C006_MOD15A2_LAI_FPAR_zmaw/tidy/",
  file_whc = "~/data/mct_data/cwdx80_forcing_halfdeg.nc",
  dir_whc_tidy = "~/data/mct_data/tidy/",
  file_landmask = "~/data/archive_legacy/landmasks/WFDEI-elevation.nc",
  dir_landmask_tidy = "~/data/archive_legacy/landmasks/WFDEI-elevation_tidy/",
  file_elv = "~/data/archive_legacy/landmasks/WFDEI-elevation.nc",
  dir_elv_tidy = "~/data/archive_legacy/landmasks/WFDEI-elevation_tidy/",
  save_drivers = TRUE,
  dir_drivers = here::here("input/tidy/"),
  overwrite = FALSE,
  spinupyears = 10,
  recycle = 1,
  dir_out = here::here("output/tidy/"),
  dir_out_nc = "xxx",  # xxx not yet handled
  save_nc = list(
    gpp = "mon"
  ),
  nthreads = 1,
  ncores_max = 1
)
```

## Specify model parameters

```{r}
par <- list(
  kphio              = 0.04998,    # setup ORG in Stocker et al. 2020 GMD
  kphio_par_a        = 0.0,        # set to zero to disable temperature-dependence of kphio
  kphio_par_b        = 1.0,
  soilm_thetastar    = 0.6 * 240,  # to recover old setup with soil moisture stress
  soilm_betao        = 0.0,
  beta_unitcostratio = 146.0,
  rd_to_vcmax        = 0.014,      # value from Atkin et al. 2015 for C3 herbaceous
  tau_acclim         = 30.0,
  kc_jmax            = 0.41
)
```

## Make forcing tidy

Convert forcing data, provided as NetCDF, into a tidy format (time series data frames for each gridcell along rows) and saved for longitudinal bands in a binary format.

```{r eval=FALSE}
# Convert forcing files to a tidy format
# Checks by file. If tidy already, skips automatically
settings <- grsofun_tidy(settings)
```

## Run model

Read forcing data and construct the driver object for an rsofun run. Run the model and save outputs in a tidy format in `.rds` files for each longitudinal band.

```{r eval=FALSE}
# Run rsofun
# Parallelizes runs to chunks of longitudinal bands
error <- grsofun_run(par, settings)
```

## Collect outputs

Read outputs saved in separate files for each longitudinal band. Collect data from all longitudinal bands and apply temporal aggregation steps as specified by the settings.

This can either write the aggregated data again to tidy files by longitudinal band, or return the temporally aggregated data into the R environment.

```{r eval=TRUE, message=FALSE}
# Collect output
# Only variables and at aggregation level required for NetCDF output
df <- grsofun_collect(settings, return_data = TRUE)
```

## Example plot

Plot a map.
```{r}
coast <- rnaturalearth::ne_coastline(
  scale = 110, 
  returnclass = "sf"
  )

# January
gg1 <- df |> 
  filter(month == 1) |> 
  ggplot() +
  geom_raster(
    aes(x = lon, y = lat, fill = gpp),
    show.legend = TRUE
    ) +
  geom_sf(
    data = coast,
    colour = 'black',
    linewidth = 0.3
  )  +
  coord_sf(
    ylim = c(-60, 85),
    expand = FALSE
  ) +
  scale_fill_viridis_c(
    name = expression(paste("gC m"^-2, "s"^-1)),
    option = "cividis",
    limits = c(0, 15)
    ) +
  theme_void() +
  labs(
    subtitle = "January monthly mean"
  )

# July
gg2 <- df |> 
  filter(month == 7) |> 
  ggplot() +
  geom_raster(
    aes(x = lon, y = lat, fill = gpp),
    show.legend = TRUE
    ) +
  geom_sf(
    data = coast,
    colour = 'black',
    linewidth = 0.3
  )  +
  coord_sf(
    ylim = c(-60, 85),
    expand = FALSE
  ) +
  scale_fill_viridis_c(
    name = expression(paste("gC m"^-2, "s"^-1)),
    option = "cividis",
    limits = c(0, 15)
    ) +
  theme_void() +
  labs(
    subtitle = "July monthly mean"
  )

cowplot::plot_grid(
  gg1, 
  gg2,
  ncol = 1
)
```
